---
title: "writeup"
author: "Logan King"
date: "9/27/2020"
output: github_document
---

# If home field advantage exists, how much of an impact does it have on winning the world series?

Home field advantage is the edge which teams are perceived to gain when playing games at their home stadium. In this blog, we will explore the effects of home field advantage in a World Series matchup between the Atlanta Braves and New York Yankees through analytical and simulation methods.

<details>
  <summary>See code here:</summary>
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load relevant packages}
library(tidyverse)
library(data.table)
library(ggplot2)
```

```{r set working directory and initialize data}
# Get all possible outcomes
apo <- fread("C:/Users/kingl/Desktop/Projects/probability_fall20/probability-and-inference-portfolio-king-logan/assets/all-possible-world-series-outcomes.csv")
```
</details>

## Analytical Probability

Analytical probability is found by taking the entire range of possible outcomes and analyzing the probability of each outcome. For example if a fair coin is flipped three times, the possible range of outcomes is {HHH, HHT, HTH, TTH, THH, ,THT, HTT, TTT}. Each outcome has a probability of 1/8 or (.5 * .5 * .5) and the analytical probability of at least two heads can be calculated from this set by summing the probabilities of all outcomes with at least two heads. 

Much like the coin example, the analytical probability of the outcome of the World Series can be calculated. In order to calculate this probability, we first need the set of all possible outcomes of the World Series games. That is provided in a previously loaded dataset. As it turns out, there are 35 possible outcomes for each team to win the World Series. However, the likelihood of a team emerging victorious depends on several factors. 

Unlike the fair coin example, the probability of victory for the Braves and Yankees are not even. For this exercise, the Braves will have a 55% chance of winning (45% chance of losing) a single game in the series in the absence of home field advantage. In order to observe the effects of home field advantage, a 10% probability of victory bonus will be given to the home team in a matchup. Effectively, this gives the braves a `r 55*1.1`% chance of victory (`r 100 - (55*1.1)`% chance of defeat) in a single home matchup and a `r 100 - (45*1.1)`% chance of victory (`r 45*1.1`% chance of defeat) in a single road matchup.

For a series in which the sequence of game locations is {NYC, NYC, ATL, ATL, ATL, NYC, NYC}, each combination of single game location (Home or Away) and result (Win or Loss) is assigned its associated probability. Each probability in the series is multiplied together in order to find the probability of that specific series outcome across the entire set of possible World Series outcomes. By summing up the probability of each series in which the outcome is a win, we can find the analytical probability of the Braves winning the world Series under the following parameters:

* Game Locations: {NYC, NYC, ATL, ATL, ATL, NYC, NYC}
* Probability that the Braves win a single game: 0.55
* Home Field Advantage Multiplier (1.0, 1.1) 
  * we can find the effects of both no observed home field advantage and observed home field advantage

<details>
  <summary>See code here:</summary>
```{r analytical probability function and values}
analytical_ws <- function(hfi, pb, adv_mult) {
  pbh <- pb * adv_mult
  pba <- 1 - (1 - pb) * adv_mult
  # Calculate the probability of each possible outcome
  apo[, p := NA_real_] # Initialize new column in apo to store prob
  for(i in 1:nrow(apo)){
    prob_game <- rep(1, 7)
    for(j in 1:7){
      p_win <- ifelse(hfi[j], pbh, pba)
      prob_game[j] <- case_when(
          apo[i,j,with=FALSE] == "W" ~ p_win
        , apo[i,j,with=FALSE] == "L" ~ 1 - p_win
        , TRUE ~ 1
      )
    }
    apo[i, p := prod(prob_game)] # Data.table syntax
  }
  if(apo[, sum(p)] %>% near(1)) {
    return(apo[, sum(p), overall_outcome])
  }
}

analytical_adv <- analytical_ws(hfi = c(0,0,1,1,1,0,0), pb = .55, adv_mult = 1.1)$V1[1]
analytical_no_adv <- analytical_ws(hfi = c(0,0,1,1,1,0,0), pb = .55, adv_mult = 1)$V1[1]
```
</details>

We find that the analytical probability of the Braves winning the World Series when home field advantage plays a factor is `r round(analytical_adv*100,3)`% while the analytical probability of the Braves winning the World Series when home field advantage does not exist is `r round(analytical_no_adv*100,3)`%. The presence of a home field advantage actually harms the Braves, because they have less home games than the Yankees in the event that the series lasts all seven games. However, they are more likely than not to win the World Series in both instances, because their single game winning probability never dips below 50% in either scenario. 

## Simulated Probability

Much like analytical probability was calculated, we can use simulation to approximate the probability of a team winning the World Series. Simulation is different in analytical probability in that it will provide an approximation of the true probability over a given set of iterations as opposed to the true probability value like the analytical method does. The simulation method can give a good representation of the variation of the outcomes of probabilistic events which can be seen in the short term, however in the long term it approaches the true probability given by the analytical method. 

Simulation is calculated by first determining the outcome of a single series. Given the inputs of location, win probability, and home field advantage multiplier, an output of win or loss is generated for a single game. This single game operation is repeated until the series is completed by either winning or losing four games. In order to simulate the probability of the Braves winning the World Series, each individual series can be replicated 100,000 times to give a reliable approximation of the true probability. 

<details>
  <summary>See code here:</summary>
```{r function for simulation}
set.seed(100)

simulated_ws <- function(hfi, pb, adv_mult) {
  pbh <- pb * adv_mult
  pba <- 1 - (1 - pb) * adv_mult
  win_count <- 0
  for(i in 1:7) {
    if(hfi[i] == 1){
      p_win <- pbh
    } else {
      p_win <- pba
    }
    game_outcome <- rbinom(1, 1, p_win)
    win_count <- win_count + game_outcome
    if(win_count == 4 | (i - win_count) == 4) break
  }
  return(win_count == 4)
}
```

```{r simulation values}
sim_outcome_adv <- NA
for(i in 1:100000){
  sim_outcome_adv[i] <- simulated_ws(hfi = c(0,0,1,1,1,0,0), pb = .55, adv_mult = 1.1)
}
simulation_adv <- mean(sim_outcome_adv)

sim_outcome_no <- NA
for(i in 1:100000){
  sim_outcome_no[i] <- simulated_ws(hfi = c(0,0,1,1,1,0,0), pb = .55, adv_mult = 1)
}
simulation_no_adv <- mean(sim_outcome_no)
```
</details>

After running the simulation, we see that the simulated probability for the Braves to win the World Series (with the same parameters as previously established) is `r round(simulation_adv*100,3)`% with the effect of home field advantage and `r round(simulation_no_adv*100,3)`% without a home field advantage effect.

As was discussed previously, simulation will always display some amount of variation from the true probability - which was seen above in our simulation results. This variation can be explained in terms of absolute and relative error. Absolute error is the total amount that an observed probability varies from the true probability, while relative error is the absolute error as a proportion of the true probability (essentially the error in the context of the true probability). 

<details>
  <summary>See code here:</summary>
```{r error values}
ae_adv <- abs(simulation_adv-analytical_adv)
ae_no_adv <- abs(simulation_no_adv-analytical_no_adv)

re_adv <- ae_adv/analytical_adv
re_no_adv <- ae_no_adv/analytical_no_adv
```
</details>

The absolute error for the simulation with home field advantage is `r round(ae_adv,5)` while the absolute error without home field advantage is `r round(ae_no_adv,5)`. The relative error seen in the simulations for home field advantage and no home field advantage is `r round(re_adv,5)` and `r round(re_no_adv,5)`, respectively.

## Bonus

We can also explore the effect of home field advantage in relation to single game winning probability and their effect on the probability of winning a series. The below plot uses analytical probability calculations to capture these interactions by displaying varying lines for both home field advantage and no home field advantage along with a varying single game probability axis in order to see how the probability of winning a World Series (located on the other axis) is influenced. Two plots are displayed, one for a series with 4 home games and one for a series with 4 away games.

<details>
  <summary>See code here (plot code hidden by echo=FALSE in next chunk):</summary>
```{r build dataset}
pb_comparison <- expand.grid(pb = seq(0, 1, by = .02),
                             adv_mult = c(1.0,1.1),
                             ws_win_prob_h = NA_real_,
                             ws_win_prob_a = NA_real_,
                             KEEP.OUT.ATTRS = F)

for(i in 1:nrow(pb_comparison)){
  pb_comparison[i,'ws_win_prob_a'] <- analytical_ws(hfi = c(0,0,1,1,1,0,0), 
                                                    pb = pb_comparison[i,'pb'], 
                                                    adv_mult = pb_comparison[i,'adv_mult'])$V1[1]
  pb_comparison[i,'ws_win_prob_h'] <- analytical_ws(hfi = c(1,1,0,0,0,1,1),
                                                    pb = pb_comparison[i,'pb'],
                                                    adv_mult = pb_comparison[i,'adv_mult'])$V1[1]
}
```
</details>

```{r plot for bonus questions, echo=FALSE}
pb_comparison$adv_mult <- pb_comparison$adv_mult %>% as.factor()

pb_comp_long <- pb_comparison %>% 
  pivot_longer(-c(pb, adv_mult), names_to = 'location', values_to = 'win_prob')

ggplot(pb_comp_long, aes(x = pb*100, y = win_prob*100, color = adv_mult)) +
  geom_line() +
  facet_wrap(~ location, 
             labeller = labeller(location = c('ws_win_prob_a' = '4 Away Games', 'ws_win_prob_h' = '4 Home Games'))) +
  labs(title = 'The Effect of Single Game Win Probability and Home Field Advantage Effects \n on World Series Win Probability',
       x = 'Probability of Winning a Single Game',
       y = 'Probability of Winning the World Series',
       color = 'Advantage Multiplier') +
  scale_x_continuous(breaks=seq(0, 100, 25),
                     labels = function(x){paste0(x, '%')}) +
  scale_y_continuous(breaks=seq(0, 100, 25),
                     labels = function(x){paste0(x, '%')})
```

From the above plots, several conclusions can be made. When a series is primarily away games for a team, no advantage is better than a 10% home field advantage boost when the probability of winning a single game is below approximately 55%. Above that point, home field advantage becomes more beneficial. This trend is reversed when a series is primarily home games for a team. In this instance, no home field advantage is only ideal up until approximately 45% probability of winning a single game. After that point, home field advantage is more beneficial.


